Soling at the 2000 Summer Olympics

### 2020-07-28 inst/f1points.Rmd
more work

```{r,label=maximumlikelihoodcalculations,cache=TRUE}
mH2016 <- maxp(H2016)
mH2017 <- maxp(H2017)
mH2018 <- maxp(H2018)
mH2019 <- maxp(H2019)
```

```{r,label=printapiechart,echo=FALSE,fig.cap="Maximum likelihood estimates \\label{piechartstrength} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, seasons 2016-2019"}
mH2016
mH2017
mH2018
mH2019
par(mfrow=c(2,2))
par(mar = c(2,2,2,2)/2)
pie(mH2016[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2016")
pie(mH2017[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2017")
pie(mH2018[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2018")
pie(mH2019[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2019")
```

@@

```{r,label=realpointsshower,echo=FALSE}
real <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),formula1_points_systems(top)$real)
real <- round(real[seq_len(top)],2)
real <- sort(real,decreasing=TRUE)
real[] <- seq_along(real)
real
```

@@

```{r,label=zipfpointsshower,echo=FALSE}
zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),formula1_points_systems(top)$zipf)
zipf <- round(zipf[seq_len(top)],2)
zipf <- sort(zipf,decreasing=TRUE)
zipf[] <- seq_along(zipf)
zipf
```

@@

Thus these two systems agree on the first three places but fourth is
awarded to Raikkonen under the F1 system and Ricciardo under Zipf.

```{r,label=likelihoodrankingshower,echo=FALSE}
jj <- mH2017
jj <- sort(jj,decreasing=TRUE)
jj[] <- seq_along(jj)
jj
```

@@

However, the points system used is essentially arbitrary.  We could
use, for example, Zipf's law to rank the drivers: award one point to
the winner, half a point to second place, one third of a point to
third, and so on; see figure \ref{orderingbypointsandlikelihood_zipf}
in which we see generally poorer agreement between points-based ranks
and likelihood-based ranks.

@@

## Possible points systems

There are a number of plausible points systems that might be used:

\begin{itemize}
\item Equal points to the top $r$ competitors
$(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, noting that
$r=1$ corresponds to a ``winner takes all'' system
\item Linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for integer $0<r<n$
\item Exponential decay $(1,r,r^2,r^3,\ldots)$ for some fixed $r<1$
\item Zipf's law $(1,1/2,1/3,1/4,\ldots)$
\item The current Formula 1 system $(25,18,15,12,10,8,6,4,2,1)$
\item The inaugural Formula 1 system $(8,6,4,3,2)$
\end{itemize}

We can now calculate number of places agreed for each of the points systems:

@@

rownames(agreement4yearsallp) <- names(formula1_points_systems(top))
colnames(agreement4yearsallp) <- 2016:2019
```

```{r}
agreement4yearsallp
```

Observing that both the likelihood ranking and the points ranking are
random variables in this paradigm suggests a method whereby we can
objectively assess a given points system.  Using sampling techniques
we can repeatedly generate an order table \emph{in silico}, using
estimated driver strengths from the observed tables.  For each of,
say, 1000 such synthetic tables, calculate drivers' maximum likelihood
\citeauthor{plackett1975} strengths, and also their points awarded
according to any given points system.  We then compare rankings
generated by the \citeauthor{plackett1975} strengths and the points
awarded and note the degree of agreement between the two.  This
furnishes an objective assessment of the points system used.

@@

```{r,label=somesummaries}
summarytablemaker <- function(OO){
  out <- data.frame(
      means = apply(OO,1,mean),
      winner_correct = rowSums(OO>1)/ncol(OO),
      all_correct = rowSums(OO==top)/ncol(OO)
  )
  rownames(out) <- names(pointslist)
  return(out)
}
s2016 <- summarytablemaker(OO2016)
s2017 <- summarytablemaker(OO2017)
s2018 <- summarytablemaker(OO2018)
s2019 <- summarytablemaker(OO2019)
```

```{r,fig=TRUE}
typecol <- c(rep("red",2),rep("orange",10),rep("yellow",11),rep("green",1),rep("blue",7),rep("purple",7))
jj <- cbind(s2016[,1],s2017[,1],s2018[,1],s2019[,1])  # mean
matplot(jj,type="n",xlab='points system',ylab='mean correct places')
points(s2016[,1],col=typecol,pch=1)
points(s2017[,1],col=typecol,pch=2)
points(s2018[,1],col=typecol,pch=3)
points(s2019[,1],col=typecol,pch=4)
legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```

```{r,fig=TRUE}
jj <- cbind(s2016[,2],s2017[,2],s2018[,2],s2019[,2])  # prob of winner
matplot(jj,type="n",xlab='points system',ylab='probability of winner correct')
points(s2016[,2],col=typecol,pch=1)
points(s2017[,2],col=typecol,pch=2)
points(s2018[,2],col=typecol,pch=3)
points(s2019[,2],col=typecol,pch=4)
legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```

```{r,fig=TRUE}
jj <- cbind(s2016[,3],s2017[,3],s2018[,3],s2019[,3])  # total rank order correct
matplot(jj,type="n",xlab='points system',ylab='probability of total rank order correct')
points(s2016[,3],col=typecol,pch=1)
points(s2017[,3],col=typecol,pch=2)
points(s2018[,3],col=typecol,pch=3)
points(s2019[,3],col=typecol,pch=4)
legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```

```{r,makeatable}
table(real_winner_ok=OO[1,]>0,top3_winner_ok=OO[3,]>0)
```

Consider the first column of the table above which shows the number of
virtual races in which each scoring system gets the winner incorrect;
a good points system has a low value here.  Note that `top3` has by
far the lowest number, a finding which I have no explanation for.

>>>>

```{r,label=maximumlikelihoodcalculations,cache=TRUE,echo=FALSE}
mH2016 <- maxp(H2016)
mH2017 <- maxp(H2017)
mH2018 <- maxp(H2018)
mH2019 <- maxp(H2019)
```

```{r,label=printapiechart,echo=FALSE,fig.cap="Maximum likelihood estimates \\label{piechartstrength} of the strengths of the top-ranked 11 drivers in Formula 1 motor racing, seasons 2016-2019"}
par(mfrow=c(2,2))
par(mar = c(2,2,2,2)/2)
pie(mH2016[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2016")
pie(mH2017[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2017")
pie(mH2018[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2018")
pie(mH2019[c(1,6,11,2,7,3,8,4,9,5,10)],cex=0.7,main="2019")
```

@@

```{r,label=realpointsshower,echo=FALSE}
real <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),formula1_points_systems(top)$real)
real <- round(real[seq_len(top)],2)
real <- sort(real,decreasing=TRUE)
real[] <- seq_along(real)
```

@@

```{r,label=zipfpointsshower,echo=FALSE}
zipf <- ordertable2points(ranktable_to_ordertable(wikitable_to_ranktable(F1_table_2017)),formula1_points_systems(top)$zipf)
zipf <- round(zipf[seq_len(top)],2)
zipf <- sort(zipf,decreasing=TRUE)
zipf[] <- seq_along(zipf)
```

@@

Thus these two systems agree on the first three places but fourth is
awarded to Raikkonen under the F1 system and Ricciardo under Zipf.
Compare the likelihood ranking:

```{r,label=likelihoodrankingshower,echo=FALSE}
jj <- mH2017
jj <- sort(jj,decreasing=TRUE)
jj[] <- seq_along(jj)
```

@@

However, the points system used is essentially arbitrary.  We could
use, for example, Zipf's law to rank the drivers: award one point to
the winner, half a point to second place, one third of a point to
third, and so on; see figure \ref{orderingbypointsandlikelihood_zipf}
in which we see generally poorer agreement between points-based ranks
and likelihood-based ranks.  There are a number of plausible points
systems that might be used:

@@

\begin{itemize}
\item The current Formula 1 system $(25,18,15,12,10,8,6,4,2,1)$
\item The inaugural Formula 1 system $(8,6,4,3,2)$
\item Zipf's law $(1,\frac{1}{2},\frac{1}{3},\frac{1}{4},\ldots)$
\item Linear decrease $(n,n-1,n-2,\ldots,3,2,1,0)$
\item Halving system:  $(1,\frac{1}{2},\frac{1}{4},\frac{1}{8},\ldots)$
\item A ``winner takes all'' system $(1,0,\ldots)$
\end{itemize}

We note that some of these may be generalized; we might consider a
generalized linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for fixed
integer $0<r<n$; the halving system can be generalized to a geometric
distribution; and the winner takes all system can be replaced by
giving equal points to the top $r$ competitors
$(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, for some integer $r<n$.

@@

rownames(agreement4yearsallp) <- names(formula1_points_systems(top))
colnames(agreement4yearsallp) <- 2016:2019
wanted <- c(1,2,24,23,3,29)
```

```{r}
agreement4yearsallp[wanted,]
```

There is no points system that is the best for all years.  However,
observing that both the likelihood ranking and the points ranking are
random variables in this paradigm suggests a method whereby we can
objectively assess a given points system.  Using sampling techniques
we can repeatedly generate an order table \emph{in silico}, using
estimated driver strengths from the observed tables.  For each of,
say, 1000 such synthetic tables, calculate drivers' maximum likelihood
\citeauthor{plackett1975} strengths, and also their points awarded
according to any given points system.  We then compare rankings
generated by the \citeauthor{plackett1975} strengths and the points
awarded and note the degree of agreement between the two.  This
furnishes an objective assessment of the points system used.

@@

```{r,label=somesummaries,echo=FALSE}
summarytablemaker <- function(OO){
  out <- data.frame(
      means = apply(OO,1,mean),
      winner_correct = rowSums(OO>1)/ncol(OO),
      all_correct = rowSums(OO==top)/ncol(OO)
  )
  rownames(out) <- names(pointslist)
  return(out)
}
s2016 <- summarytablemaker(OO2016)
s2017 <- summarytablemaker(OO2017)
s2018 <- summarytablemaker(OO2018)
s2019 <- summarytablemaker(OO2019)
```

```{r,label=alottaplotsmeancorrect,echo=FALSE,fig=TRUE}
jj <- cbind(s2016[,1],s2017[,1],s2018[,1],s2019[,1])  # mean

#typecol <- c(rep("red",2),rep("orange",10),rep("yellow",11),rep("green",1),rep("blue",7),rep("purple",7))
#matplot(jj,type="n",xlab='points system',ylab='mean correct places')
#points(s2016[,1],col=typecol,pch=1)
#points(s2017[,1],col=typecol,pch=2)
#points(s2018[,1],col=typecol,pch=3)
#points(s2019[,1],col=typecol,pch=4)
#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))

matplot(jj[wanted,],type="n",xlab='points system',ylab='mean correct places',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2,at=1:4)
points(s2016[wanted,1],pch=1)
points(s2017[wanted,1],pch=2)
points(s2018[wanted,1],pch=3)
points(s2019[wanted,1],pch=4)
legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```

```{r,label=alottaplotswinnercorrect,echo=FALSE,fig=TRUE}
jj <- cbind(s2016[,2],s2017[,2],s2018[,2],s2019[,2])  # prob of winner
#matplot(jj,type="n",xlab='points system',ylab='probability of winner correct')
#points(s2016[,2],col=typecol,pch=1)
#points(s2017[,2],col=typecol,pch=2)
#points(s2018[,2],col=typecol,pch=3)
#points(s2019[,2],col=typecol,pch=4)
#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))

matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of winner correct',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2)
points(s2016[wanted,2],pch=1)
points(s2017[wanted,2],pch=2)
points(s2018[wanted,2],pch=3)
points(s2019[wanted,2],pch=4)
legend(x=3.4,y=0.67,lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```

```{r,label=alottaplotstotalrankorder,echo=FALSE,fig=TRUE}
jj <- cbind(s2016[,3],s2017[,3],s2018[,3],s2019[,3])  # total rank order correct
#matplot(jj,type="n",xlab='points system',ylab='probability of total rank order correct')
#points(s2016[,3],col=typecol,pch=1)
#points(s2017[,3],col=typecol,pch=2)
#points(s2018[,3],col=typecol,pch=3)
#points(s2019[,3],col=typecol,pch=4)
#legend("topright",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))

matplot(jj[wanted,],type="n",xlab='points system',ylab='probability of total rank order correct',axes=FALSE)
axis(1,at=1:6,labels=names(formula1_points_systems(top)[wanted]))
axis(2)
points(s2016[wanted,3],pch=1)
points(s2017[wanted,3],pch=2)
points(s2018[wanted,3],pch=3)
points(s2019[wanted,3],pch=4)
legend("topleft",lty=0,pch=1:4,legend=c("2016","2017","2018","2019"))
```
